Comparison among doublet predictions on benchmarking datasets

### 2024-06-26 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
need to use purrr::map() to get patchwork to print. results in extra names printing out, but it is what it is probably
Now making the correct PCA plots with single function, and added section for comparing 2 methods

```{r}
plot_pca_calls <- function(df, 
                           color_column, 
                           pred_column,
                           title) {
  # Plot PCs colored by singlet/doublet, showing doublets on top
  # df is expected to contain columns PC1, PC2, `color_column`, and `pred_column`. These should _not_ be provided as strings
  ggplot(df) + 
    aes(x = PC1, 
        y = PC2, 
        color = {{color_column}}) +
  geom_point(
    size = 0.75, 
    alpha = 0.6
  ) +
  scale_color_manual(name = "", values = c("black", "gray80")) + 
  geom_point(
    data = dplyr::filter(df, {{color_column}} == "doublet"), 
    color = "black",
    size = 0.75
  ) +
  ggtitle(title) 
}

plot_pca_metrics <- function(df, color_column, metric_colors) {
  # Plot PCs colored by performance metric, showing false calls on top
  # metric_colors is a named vector of colors used for coloring tp/tn/fp/fn
  # df is expected to contain columns PC1, PC2, and `color_column`. This should _not_ be provided as a string.
  ggplot(df) + 
    aes(x = PC1, 
        y = PC2, 
        color = {{color_column}}) +
  geom_point(
    size = 0.75, 
    alpha = 0.6
  ) + 
  geom_point(
    data = dplyr::filter(df, {{color_column}} %in% c("fp", "fn")), 
    size = 0.75
  ) +
  scale_color_manual(name = "Call type", values = metric_colors) +
  ggtitle("Consensus class metrics")
}
```

## Read and prepare input data

@@

```{r read_data}
# used in PCA plots
confusion_colors <- c(
  "tp" = "lightblue",
  "tn" = "pink",
  "fp" = "blue",
  "fn" = "firebrick2"
)

# Read in data for analysis
doublet_df_list <- dataset_names |>
  purrr::map(
    \(dataset) {

@@

For the first five panels, doublets are shown in black and singlets in light gray.

```{r, fig.width = 12, fig.height = 6}
doublet_df_list |>
  purrr::iwalk(
    \(df, dataset) {

@@

      pc_plot_list <- pc_color_columns |> 
        purrr::imap(
          \(color_column, plot_title) {
            plot_pca_calls(
              df, 
              color_column = !!sym(color_column), 
              title = plot_title
            )}
        )

      # note that the list name isn't used here
      pc_plot_list$metrics <- plot_pca_metrics(
        df,
        confusion_call, 
        metric_colors = confusion_colors
      )

@@

>>>>

```{r}
plot_pca_calls <- function(df, 
                           color_column, 
                           plot_type = "calls",
                           title) {
  # Plot PCs colored by either:
  # if plot_type == "calls", color by singlet or doublet, showing doublets on top
  # if plot_type == "class", color by singlet, doublet, or ambiguous, showing doublets & ambiguous on top
  # df is expected to contain columns PC1, PC2, `color_column`, which should _not_ be provided as a string
  plot <- ggplot(df) + 
    aes(
      x = PC1, 
      y = PC2, 
      color = {{color_column}}
    ) +
    geom_point(
      size = 0.75, 
      alpha = 0.6
    ) +
    ggtitle(title)

  if (plot_type == "calls") {
    plot <- plot + 
      scale_color_manual(
        name = "", 
        values = c("doublet" = "black", "singlet" = "gray90")) + 
      geom_point(
        data = dplyr::filter(df, {{color_column}} == "doublet"), 
        color = "black",
        size = 0.75
      )
  } else if (plot_type == "class") {
    plot <- plot + 
      scale_color_manual(
        name = "", 
        values = c("ambiguous" = "lightblue", 
                   "doublet" = "black", 
                   "singlet" = "gray90")
      ) + 
      geom_point(
        data = dplyr::filter(df, consensus_class != "singlet"), 
        size = 0.75,
        alpha = 0.7,
        aes(color = consensus_class),
      )
  } else {
    stop("plot_type must be 'calls' or 'class'")
  }

  return(plot)
}
```

## Explore consensus across methods

### Read and prepare input data

@@

```{r read_data}
# Read in data for analysis
doublet_df_list <- dataset_names |>
  purrr::map(
    \(dataset) {

@@

- For the first five panels, doublets are shown in black and singlets in light gray.
- In the sixth panel showing the consensus class, doublets are shown in black, singlets in light gray, and ambiguous are blue.

```{r, fig.width = 12, fig.height = 6}
doublet_df_list |>
  purrr::imap(
    \(df, dataset) {

@@

      pc_plot_list <- pc_color_columns |> 
        purrr::imap(
          \(color_column, plot_title) {
            plot_pca_calls(
              df, 
              color_column = !!sym(color_column), 
              plot_type = "calls",
              title = plot_title
            )}
        ) 
      # plot for consensus_class, which needs three colors
      # note that the list name isn't actually used here
      pc_plot_list$consensus_class <- plot_pca_calls(
        df, 
        color_column = consensus_class,
        plot_type = "class",
        title = "Consensus class"
      )

@@

## Explore consensus of `scDblFinder` and `cxds`

The above analysis suggests that `scrublet` may be more conservative than the other two methods, so including it in a consensus may be increasing the number of false positives.
Here, we'll repeat the same consensus analyses, but considering _only_ `scDblFinder` and `cxds`.
Since there are only two methods here, we will create two columns for this data:

### Prepare data

```{r}
# modify existing consensus columns to only consider 2 methods
doublet_df_list_2 <- doublet_df_list |>
  purrr::map(
    \(dataset_df) {

      dataset_df <- dataset_df |>
        dplyr::rowwise() |>
        # Add column `consensus_call`
        dplyr::mutate(
          consensus_call = dplyr::if_else(
            all(c(scdbl_prediction, cxds_prediction) == "doublet"),
            "doublet", 
            "singlet"
          )
        ) |>
        # Add column `consensus_class`
        dplyr::mutate(
          consensus_class = dplyr::case_when(
            consensus_call == "doublet" ~ "doublet",
            all(c(scdbl_prediction, cxds_prediction) == "singlet") ~ "singlet", 
            .default = "ambiguous"
          )
         ) |>
        # Add column `confusion_call`
        dplyr::mutate(
          confusion_call = dplyr::case_when(
            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
          )
        )

1. `scDblFinder` singlet/doublet calls
2. `cxds` singlet/doublet calls
3. Ground truth single/doublets
4. Consensus _call_ droplets are either "doublet" or "singlet"
5. Consensus _class_: droplets are either "doublet", "singlet", or "ambiguous"

For the first four panels, doublets are shown in black and singlets in light gray.

```{r, fig.width = 12, fig.height = 6}
doublet_df_list_2 |>
  purrr::imap(
    \(df, dataset) {

      pc_color_columns <- c("scDblFinder prediction" = "scdbl_prediction",
                            "cxds prediction" = "cxds_prediction",
                            "Ground truth" = "ground_truth", 
                            "Consensus call" = "consensus_call")

      pc_plot_list <- pc_color_columns |> 
        purrr::imap(
          \(color_column, plot_title) {
            plot_pca_calls(
              df, 
              color_column = !!sym(color_column), 
              plot_type = "calls",
              title = plot_title
            )}
        ) 
      # plot for consensus_class, which needs three colors
      # note that the list name isn't actually used here
      pc_plot_list$consensus_class <- plot_pca_calls(
        df, 
        color_column = consensus_class,
        plot_type = "class",
        title = "Consensus class"
      )

```{r}
metric_df <- doublet_df_list_2 |>
  purrr::imap( 
    \(df, dataset) {
        print(glue::glue("======================== {dataset} ========================"))
